Popularity Adjusted Block Models are Generalized Random Dot Product Graphs

### 2021-09-01 summary.tex
syncing what i have so far

To complete the proof of Theorem~\ref{theorem3}, recall that \(X\) is block diagonal with each block
corresponding to one community and hence \(X (X^\top X)^{-1} X^\top\) is also a
block diagonal matrix with each block corresponding to a community. 
As $B = n VV^{\top} = n X (X^\top X)^{-1} X^\top$, 
we conclude that $B_{ij} = 0$ 
whenever vertices $i$ and $j$ belong to different communities.  
\end{proof}

Theorem \ref{theorem3} provides perfect community detection given \(P\).
Letting \(|B|\) be the affinity matrix for graph \(G'\), \(G'\) is
partitioned into at least \(K\) disjoint subgraphs because each of the
\(K\) communities have no edges between them. Similar to the subspace
detection property, it could be the case that some of the communities
are represented by multiple disjoint subgraphs in \(G'\), in which case
additional reconstruction is required to identify the communities
exactly.

Using \(A\) instead of \(P\) to estimate $B$ introduces error, 
which converges to \(0\) almost surely. 
This is shown via properties of the GRDPG and ASE. 
We use this fact to motivate Orthogonal Spectral Clustering (Alg. 1). 
In order to show that OSC is consistent, 
we introduce the sparsity factor $\rho_n \in (0, 1]$, 
which can be characterized as the average expected degree of the vertices 
divided by the number of vertices. 
The treatment is the same as in \citet{307cbeb9b1be48299388437423d94bf1} (PABM) 
and \citet{rubindelanchy2017statistical} (GRDPG) in which the popularity vectors 
are scaled as $\rho_n^{1/2} \lambda^{(k \ell)}$. 
This results in the edge probability matrix being scaled as 
$P_{ij} = \rho_n \lambda_{i z_j} \lambda_{j z_i}$. 
Since $\rho_n$ is unidentifiable, we use the convention that the popularity parameters are drawn such that $\|\Lambda\|_F = O(\sqrt{n})$ and require that the $L^2$ norm of each row is bounded away from 0 and 1.

@@

For comparing methods, we define the community detection error as:
\[L_c(\hat{\sigma}, \sigma; \{v_i\}) =
\min_\pi \sum_i I(\pi \circ \hat{\sigma}(v_i) = \sigma(v_i))\]
where \(\sigma(v_i)\) is the true community label of vertex \(v_i\),
\(\hat{\sigma}(v_i)\) is the predicted label of \(v_i\), and \(\pi\) is
a permutation operator. This is effectively the ``misclustering count''
of clustering function \(\hat{\sigma}\).

For parameter estimation, because the popularity parameters $\{\lambda_{ik}\}$ are unidentifiable, we instead estimate the edge probabilities $p_{ij} = \lambda_{i z_j} \lambda_{j, z_i}$. The parameter estimation error is the root mean square error between the true and estimated edge probabilities, which is equivalent to the Frobenius norm of $P$ divided by the number of vertices:

>>>>

To complete the proof of Theorem~\ref{theorem3}, recall that \(X\) 
is block diagonal with each block corresponding to one community, 
and hence \(X (X^\top X)^{-1} X^\top\) is also a
block diagonal matrix with each block corresponding to a community. 
As $B = n VV^{\top} = n X (X^\top X)^{-1} X^\top$, 
we conclude that $B_{ij} = 0$ 
whenever vertices $i$ and $j$ belong to different communities.  
\end{proof}

Under the condition that $\|\Lambda\|_F > 0$, where 
$\Lambda$ is the $n \times K$ matrix of popularity parameters, 
Theorem \ref{theorem3} provides perfect community detection from \(P\).
Letting \(|B|\) be the affinity matrix for graph \(G'\), 
\(G'\) is partitioned into \(K\) disjoint subgraphs, as each of the
\(K\) communities have no edges between them. 
Then mapping each subgraph to a community provides a clustering with zero error 
(up to permutation).

In practice, we observe $A$ instead of $P$. 
Using \(A\) instead of \(P\) to estimate $B$ as in Theorem \ref{theorem3} introduces error, 
which converges to \(0\) almost surely. 
This is shown via properties of the GRDPG and ASE. 
We use this fact to motivate Orthogonal Spectral Clustering (Alg. 1). 
In order to show that OSC is consistent, 
we introduce the sparsity factor $\rho_n \in (0, 1]$, 
which can be characterized as the average expected degree of the vertices 
divided by the number of vertices. 
The treatment is the same as in \citet{307cbeb9b1be48299388437423d94bf1} (PABM) 
and \citet{rubindelanchy2017statistical} (GRDPG) in which the popularity vectors 
are scaled as $\rho_n^{1/2} \lambda^{(k \ell)}$. 
This results in the edge probability matrix being scaled as 
$P_{ij} = \rho_n \lambda_{i z_j} \lambda_{j z_i}$. 
Since $\rho_n$ is unidentifiable, we use the convention that the popularity parameters are drawn such that $\|\Lambda\|_F = O(\sqrt{n})$ and require that the $L^2$ norm of each row is bounded away from 0 and 1.

@@

For comparing methods, we define the community detection error as:
\[L_c(\hat{\sigma}, \sigma; V) =
\min_\pi \sum_i I(\pi \circ \hat{\sigma}(v_i) = \sigma(v_i))\]
where \(\sigma(v_i)\) is the true community label of vertex \(v_i\),
\(\hat{\sigma}(v_i)\) is the predicted label of \(v_i\), and \(\pi\) is
a permutation operator. This is effectively the ``misclustering count''
of clustering function \(\hat{\sigma}\).

For parameter estimation, because the popularity parameters $\{\lambda_{ik}\}$ are unidentifiable, we instead estimate the edge probabilities $P_{ij} = \lambda_{i z_j} \lambda_{j, z_i}$. The parameter estimation error is the root mean square error between the true and estimated edge probabilities, which is equivalent to the Frobenius norm of $P$ divided by the number of vertices:
